Analytic Doppler Imaging

### 2019-08-13 tex/ms.tex
more words

\subsection{Solution for the intensity}
\label{sec:solve_y}
For a fixed, spatially constant rest-frame spectrum $\bvec{s}$, the model
for the flux vector may be written as
%
\begin{align}
    \label{eq:inverse:fy}
    \bvec{f}
    &=
    \Doppler
    \,
    \bvec{S}
    \,
    \bvec{y}
    \quad,
\end{align}
%
where
%
\begin{align}
    \setstackgap{L}{1.25\baselineskip}
    \bvec{S} =
        \begin{pmatrix}
        \quadquad\bvec{s}\quadquad & &            &            &  \\
        & \quadquad\bvec{s}\quadquad &            &            &  \\
        &            & \quadquad\bvec{s}\quadquad &            &  \\
        &            &            & \quadquad\bvec{s}\quadquad &  \\
        &            &            &               & \ddots
        \end{pmatrix}
\end{align}
%
is a $(NK' \times N)$ block matrix constructed by repeating the column
vector $\bvec{s}$ $N$ times in blocks along the main diagonal. Given
prior mean $\boldsymbol{\mu}_\mathbf{y}$ and covariance
$\boldsymbol{\Lambda}_\mathbf{y}$ on $\mathbf{y}$, the MAP solution
is given by
%
\begin{align}
    \label{eq:inverse:yhat}
    \bvec{\hat{y}} &= 
    \boldsymbol{\Sigma}_\mathbf{\hat{y}}
    \left(
        \bvec{S}^\top\Doppler^\top
        {\boldsymbol{\Sigma}_\mathbf{f}}^{-1}
        \bvec{f}
        +
        {\boldsymbol{\Lambda}_\mathbf{y}}^{-1} \boldsymbol{\mu}_\mathbf{y}
    \right)
    \quad,
\end{align}
%
with covariance
%
\begin{align}
    \label{eq:inverse:ycov}
    \boldsymbol{\Sigma}_\mathbf{\hat{y}} &= 
    \left(
        \bvec{S}^\top\Doppler^\top
        {\boldsymbol{\Sigma}_\mathbf{f}}^{-1}
        \Doppler\bvec{S}
        +
        {\boldsymbol{\Lambda}_\mathbf{y}}^{-1}
    \right)^{-1}
    \quad.
\end{align}
%
If we choose a prior with zero mean ($\boldsymbol{\mu}_\bvec{y} = 0$)
and a diagonal covariance that is constant for each spherical harmonic 
order $m$,
%
\begin{align}
    \boldsymbol{\Lambda}_\bvec{y} &=
    \mathrm{diag} \left(
        \sigma_0^2
        \quad\quad\quad\quad
        \sigma_1^2
        \quad\quad\quad\quad
        \sigma_1^2
        \quad\quad\quad\quad
        \sigma_1^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \cdots
    \right)
    \quad,
\end{align}
%
we are in fact imposing an isotropic prior with an angular power spectrum
given by $C_l = \sigma_l^2$
\citep[e.g.,][]{Baldi2006}. This is especially useful if the typical
angular scale of features on the surface of the star is known or if
a Gaussian prior can be placed on it.

@@

\subsection{The full solution}
In the previous two sections we showed how, if the stellar
rest-frame spectrum is spatially constant and known \emph{a priori} 
(from, say, a template spectrum), the posterior distribution
for the surface map $\bvec{y}$ is \emph{analytic} 
(Equations~\ref{eq:inverse:yhat}
and \ref{eq:inverse:ycov}). Conversely, if the surface map is known,
the posterior for the spectrum is also analytic 
(Equations~\ref{eq:inverse:shat} and \ref{eq:inverse:scov}). Both sets of
solutions require Gaussian priors on the quantities of interest, but these
are easily interpretable. A Gaussian prior on
the spherical harmonic coefficients of the surface map corresponds 
to the angular power spectrum of the surface features, whereas a Gaussian
prior on the spectrum can be used to correctly propagate uncertainties in
the stellar spectrum from observational uncertainties on the stellar
metallicity, temperature, etc.

@@

%
\begin{align}
    \label{eq:norm:f}
    \bvec{f}
    &=
    \frac{
        \Doppler
        \,
        \azero
    } {
        \bvec{b}
    }
    \quad,
\end{align}
%
where $\bvec{b}$ is the baseline and the vector division is
taken elementwise. When $\azero$ is given by Equation~(\ref{eq:inverse:azero}),
the baseline $\bvec{b}$ is simply the total (wavelength-integrated) flux
at each epoch, which \citet{Luger2019} showed is a linear operation on the
spherical harmonic vector $\bvec{y}$:
%
\begin{align}
    \label{eq:norm:b}
    \bvec{b}
    &=
    \bvec{r}^\top \bvec{A_1} \bvec{R}(t) \, \bvec{y}
    \nonumber \\
    &\equiv
    \bvec{B} \, \bvec{y}
\end{align}
%
where $\bvec{r}$ is given by Equation~(18) in \citet{Luger2019},
$\bvec{A_1}$ is given in Appendix B of \citet{Luger2019}, and \bvec{R}(t)
is the rotation matrix defined previously.

Given this new equation, we must change how we approach the inverse problem
(\S\ref{sec:inverse}). When solving for the spectrum given the spherical 
harmonic vector (\S\ref{sec:solve_s}), we need only make the replacements
$\bvec{f} \rightarrow \nicefrac{\bvec{f}}{\bvec{B}\, \bvec{y}}$
and 
$\boldsymbol{\Sigma}_\mathbf{f} \rightarrow \nicefrac{\boldsymbol{\Sigma}_\mathbf{f}}{\bvec{B}\, \bvec{y}}$ 
in Equations~(\ref{eq:inverse:shat}) and (\ref{eq:inverse:scov}).
However, when solving for the spherical harmonic vector given the spectrum
(\ref{sec:solve_y}), the problem is no longer linear in $\bvec{y}$, as the
vector of interest appears in both the numerator and the denominator:
%
\begin{align}
    \label{eq:norm:fy}
    \bvec{f}
    &=
    \frac{
        \Doppler
        \,
        \bvec{S}
        \,
        \bvec{y}
    }{
        \bvec{B}
        \,
        \bvec{y}
    }
    \quad.
\end{align}
%
While we can solver for $\bvec{y}$ using a non-linear optimizer, it is
often the case that the denominator in Equation~(\ref{eq:norm:fy}) is close
to unity; i.e., the total flux from the star typically varies by only a
few percent. When this is the case, it is straightforward to linearize
this equation.

>>>>

\subsection{Solution for the intensity}
\label{sec:solve_y}
For a fixed, spatially constant rest-frame spectrum $\bvec{s}$, the model
for the flux vector may be written as
%
\begin{align}
    \label{eq:inverse:fy}
    \bvec{f}
    &=
    \Doppler
    \,
    \bvec{S}
    \,
    \bvec{y}
    \quad,
\end{align}
%
where
%
\begin{align}
    \setstackgap{L}{1.25\baselineskip}
    \bvec{S} =
        \begin{pmatrix}
        \quadquad\bvec{s}\quadquad & &            &            &  \\
        & \quadquad\bvec{s}\quadquad &            &            &  \\
        &            & \quadquad\bvec{s}\quadquad &            &  \\
        &            &            & \quadquad\bvec{s}\quadquad &  \\
        &            &            &               & \ddots
        \end{pmatrix}
\end{align}
%
is a $(NK' \times N)$ block matrix constructed by repeating the column
vector $\bvec{s}$ $N$ times in blocks along the main diagonal. 
%
Before we solve this equation for $\bvec{y}$, we note that in practice
it is convenient to fix the coefficient of the $Y_{0,0}$ spherical
harmonic, $y_0$, to unity. After all, this coefficient is not directly
constrained by the data, as it merely sets the baseline intensity on the
surface of the star and is therefore formally degenerate with the stellar 
luminosity (which also cannot be directly inferred from the data). This
choice further ensures a mean baseline of unity for the model 
vector $\mathbf{f}$. We may therefore write
%
\begin{align}
    \label{eq:inverse:fy1}
    \bvec{f}
    &=
    \Doppler
    \,
    \bvec{s}_\bvec{0}
    \,
    +
    \,
    \Doppler
    \,
    \bvec{S_1}
    \,
    \bvec{y}_\bvec{1}
    \quad,
\end{align}
%
where the vector $\bvec{s_0}$ denotes the first column of $\bvec{S}$,
the matrix $\bvec{S_1}$ denotes the remaining columns of $\bvec{S}$, and
$\bvec{y_1}$ is the quantity we wish to solve for: the vector of spherical 
harmonic coefficients of degree $l \ge 1$.

Given prior mean $\boldsymbol{\mu}_\bvec{y_1}$ and covariance
$\boldsymbol{\Lambda}_\bvec{y_1}$ on $\bvec{y_1}$, the MAP solution
is given by
%
\begin{align}
    \label{eq:inverse:y1hat}
    \bvec{\hat{y}_1} &= 
    \boldsymbol{\Sigma}_\mathbf{\hat{y}_1}
    \left(
        \bvec{S_1}^\top\Doppler^\top
        {\boldsymbol{\Sigma}_\mathbf{f}}^{-1}
        \left(\bvec{f} - \Doppler\bvec{s}_\bvec{0}\right)
        +
        {\boldsymbol{\Lambda}_\bvec{y_1}}^{-1} \boldsymbol{\mu}_\bvec{y_1}
    \right)
    \quad,
\end{align}
%
with covariance
%
\begin{align}
    \label{eq:inverse:y1cov}
    \boldsymbol{\Sigma}_\bvec{\hat{y}_1} &= 
    \left(
        \bvec{S_1}^\top\Doppler^\top
        {\boldsymbol{\Sigma}_\bvec{f}}^{-1}
        \Doppler\bvec{S_1}
        +
        {\boldsymbol{\Lambda}_\bvec{y_1}}^{-1}
    \right)^{-1}
    \quad.
\end{align}
%
If we choose a prior with zero mean ($\boldsymbol{\mu}_\bvec{y_1} = 0$)
and a diagonal covariance that is constant for each spherical harmonic 
order $m$,
%
\begin{align}
    \boldsymbol{\Lambda}_\bvec{y_1} &=
    \mathrm{diag} \left(
        \sigma_1^2
        \quad\quad\quad\quad
        \sigma_1^2
        \quad\quad\quad\quad
        \sigma_1^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \sigma_2^2
        \quad\quad\quad\quad
        \cdots
    \right)
    \quad,
\end{align}
%
we are in fact imposing an isotropic prior with an angular power spectrum
given by $C_l = \sigma_l^2$
\citep[e.g.,][]{Baldi2006}. This is especially useful if the typical
angular scale of features on the surface of the star is known or if
a Gaussian prior can be placed on it.

@@

\subsection{The full solution}
In the previous two sections we showed how, if the stellar
rest-frame spectrum is spatially constant and known \emph{a priori} 
(from, say, a template spectrum), the posterior distribution
for the surface map $\bvec{y}$ is \emph{analytic} 
(Equations~\ref{eq:inverse:y1hat}
and \ref{eq:inverse:y1cov}). Conversely, if the surface map is known,
the posterior for the spectrum is also analytic 
(Equations~\ref{eq:inverse:shat} and \ref{eq:inverse:scov}). Both sets of
solutions require Gaussian priors on the quantities of interest, but these
are easily interpretable. A Gaussian prior on
the spherical harmonic coefficients of the surface map corresponds 
to the angular power spectrum of the surface features, whereas a Gaussian
prior on the spectrum can be used to correctly propagate uncertainties in
the stellar spectrum from observational uncertainties on the stellar
metallicity, temperature, etc.

@@

%
\begin{align}
    \label{eq:norm:f}
    \bvec{f}
    &=
    \frac{
        \Doppler
        \,
        \azero
    } {
        \bvec{b}
    }
    \quad,
\end{align}
%
where $\bvec{b}$ is the baseline and the vector division is
taken elementwise. When $\azero$ is given by Equation~(\ref{eq:inverse:azero}),
the baseline $\bvec{b}$ is simply the total (wavelength-integrated) flux
at each epoch, repeated for every element in each spectrum. 
Based on \citet{Luger2019}, this can be computed as a purely linear operation 
on the spherical harmonic vector $\bvec{y}$:
%
\begin{align}
    \label{eq:norm:b}
    \bvec{b}
    &=
    %
    \begin{pmatrix}
        \bvec{r}^\top \bvec{A_1} \bvec{R}(t_0) \\
        \cdots \\
        \bvec{r}^\top \bvec{A_1} \bvec{R}(t_0) \\
        \bvec{r}^\top \bvec{A_1} \bvec{R}(t_1) \\
        \cdots \\
        \bvec{r}^\top \bvec{A_1} \bvec{R}(t_1) \\
        \cdots \\
        \bvec{r}^\top \bvec{A_1} \bvec{R}(t_{M-1}) \\
        \cdots \\
        \bvec{r}^\top \bvec{A_1} \bvec{R}(t_{M-1})
    \end{pmatrix}
    \bvec{y}
    %
    \nonumber \\[0.5em]
    &\equiv
    \bvec{B} \, \bvec{y}
\end{align}
%
where $\bvec{r}$ is given by Equation~(18) in \citet{Luger2019},
$\bvec{A_1}$ is given in Appendix B of \citet{Luger2019}, and $\bvec{R}(t)$
is the rotation matrix defined previously.

Given this new equation, we must change how we approach the inverse problem
(\S\ref{sec:inverse}). When solving for the spectrum given the spherical 
harmonic vector (\S\ref{sec:solve_s}), we need only make the replacements
$\bvec{f} \rightarrow \nicefrac{\bvec{f}}{\bvec{B}\, \bvec{y}}$
and 
$\boldsymbol{\Sigma}_\mathbf{f} \rightarrow \nicefrac{\boldsymbol{\Sigma}_\mathbf{f}}{\bvec{B}\, \bvec{y}}$ 
in Equations~(\ref{eq:inverse:shat}) and (\ref{eq:inverse:scov}).
However, when solving for the spherical harmonic vector given the spectrum
(\ref{sec:solve_y}), the problem is no longer linear in $\bvec{y}$, as the
vector of interest appears in both the numerator and the denominator:
%
\begin{align}
    \label{eq:norm:fy}
    \bvec{f}
    &=
    \frac{
        \bvec{D}
        \,
        \bvec{S}
        \,
        \bvec{y}
    }{
        \bvec{B}
        \,
        \bvec{y}
    }
    \quad.
\end{align}
%
While we can solve for $\bvec{y}$ using a non-linear optimizer, it is
often the case that the denominator in Equation~(\ref{eq:norm:fy}) is close
to unity; i.e., the total flux from the star typically varies by only a
few percent. When this is the case, it is straightforward to linearize
this equation.
Following the notation from \S\ref{sec:solve_y}, we may therefore write
%
\begin{align}
    \bvec{f}
    &=
    \frac{
        \bvec{D} \, \bvec{s_0}
        \,
        + 
        \,
        \bvec{D}\bvec{S_1}
        \,
        \bvec{y_1}
    }{
        \bvec{B_0}
        \,
        +
        \bvec{B_1}
        \,
        \bvec{y_1}
    }
    \quad,
\end{align}
%
where $\bvec{B_0} = \bvec{1}$ is the first column of $\bvec{B}$
and $\bvec{B}_\bvec{1}$ denotes the
matrix composed from the remaining $N - 1$ columns of $\bvec{B}$.
Expanding the expression above about 
$\bvec{B_1} \bvec{y_1} = \bvec{0}$, we have
%
\begin{align}
    \bvec{f}
    &\approx
    \left(
        \bvec{D} \, \bvec{s_0}
        \,
        + 
        \,
        \bvec{D}\bvec{S_1}
        \,
        \bvec{y_1}
    \right)
    \circ
    \left(
        \bvec{1}
        \,
        -
        \bvec{B_1}
        \,
        \bvec{y_1}
    \right)
    \nonumber \\
    &=
    \left(
        \bvec{D}\bvec{S_1}
        \,
        -
        \bvec{C}
    \right)
    \bvec{y_1}
    \,
    +
    \,
    \bvec{D} \, \bvec{s_0}
    \quad,
\end{align}
%
where $\circ$ denotes an elementwise product and $\bvec{C}$ is a matrix
constructed by multiplying the $i^\mathrm{th}$ row of $\bvec{B_1}$ by the
$i^\mathrm{th}$ element of $\bvec{D}\bvec{s_0}$.
%
This equation is now
linear in $\bvec{y}_\bvec{1}$, so we may solve the least-squares problem
as before:
%
\begin{align}
    \label{eq:norm:yhat}
    \bvec{\hat{y}_1} &= 
    \boldsymbol{\Sigma}_\mathbf{\hat{y}_1}
    \left(
        \left(\bvec{D}\bvec{S_1} - \bvec{C}\right)^\top
        {\boldsymbol{\Sigma}_\mathbf{f}}^{-1}
        (\bvec{f} - \bvec{D}\bvec{s_0})
        +
        {\boldsymbol{\Lambda}_{\mathbf{y}_\bvec{1}}}^{-1} \boldsymbol{\mu}_{\mathbf{y}_\bvec{1}}
    \right)
    \quad,
\end{align}
%
with covariance
%
\begin{align}
    \label{eq:norm:ycov}
    \boldsymbol{\Sigma}_\mathbf{\hat{y}_1} &= 
    \left(
        \left(\bvec{D}\bvec{S_1} - \bvec{C}\right)^\top
        {\boldsymbol{\Sigma}_\mathbf{f}}^{-1}
        \left(\bvec{D}\bvec{S_1} - \bvec{C}\right)
        +
        {\boldsymbol{\Lambda}_{\mathbf{y}_1}}^{-1}
    \right)^{-1}
    \quad.
\end{align}
